---
title: "Assignment 5B - Chess ELO Calculations"
author: "Akash Sourav"
date: "2026-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Approach

I will build on the chess tournament data from Project 1. For each player, I will calculate an Elo expected score against each actual opponent using both players' pre-tournament ratings. I will add these expected scores, subtract the total from the player's actual tournament score, and identify the five largest positive and negative differences.

## Data Import

```{r data extraction}
tournament <- readLines("tournamentinfo.txt")
head(tournament, 10)
```

## Data Parsing

```{r clean up data}
player_lines <- tournament[seq(5, length(tournament), by = 3)]

head(player_lines)
length(player_lines)
```



```{r extract all player info}
player_number <- as.numeric(
  trimws(substr(player_lines, 1, 5))
)

player_name <- trimws(
  substr(player_lines, 7, 40)
)

total_points <- as.numeric(
  trimws(substr(player_lines, 42, 45))
)

players <- data.frame(
  player_number,
  player_name,
  total_points
)

head(players)
```


```{r add rating colomn}
rating_lines <- tournament[seq(6, length(tournament), by = 3)]

players$state <- trimws(substr(rating_lines, 1, 5))

players$pre_rating <- as.numeric(
  sub(".*R:\\s*([0-9]+).*", "\\1", rating_lines)
)

head(players)
```


```{r}
round_results <- strsplit(player_lines, "|", fixed = TRUE)

opponent_numbers <- lapply(round_results, function(fields) {
  as.numeric(stringr::str_extract(fields[4:10], "\\d+"))
})

opponent_numbers[[1]]
```



```{r}
players$avg_opponent_rating <- sapply(opponent_numbers, function(opponents) {
  opponent_ratings <- players$pre_rating[
    match(opponents, players$player_number)
  ]

  round(mean(opponent_ratings, na.rm = TRUE))
})

head(players)
```

## Elo Expected Scores

I use the expected-score formula from Cedric Beaume's [Understanding the Elo rating system, equation 32](https://cbeaume.com/download/EloRatingSystem.pdf):

\[
E_A = \frac{1}{1 + 10^{(R_B - R_A)/400}}
\]

Here, \(R_A\) is the player's pre-tournament rating and \(R_B\) is the opponent's pre-tournament rating. An expected score includes wins and draws: equal ratings give an expected score of 0.5 points.

```{r expected scores}
expected_scores <- lapply(seq_len(nrow(players)), function(i) {
  opponents <- opponent_numbers[[i]]
  opponents <- opponents[!is.na(opponents)]

  opponent_ratings <- players$pre_rating[
    match(opponents, players$player_number)
  ]

  # Missing ratings for actual opponents should not be silently ignored.
  stopifnot(!is.na(players$pre_rating[i]), !anyNA(opponent_ratings))

  1 / (1 + 10^((opponent_ratings - players$pre_rating[i]) / 400))
})

players$expected_score <- sapply(expected_scores, sum)
players$difference <- players$total_points - players$expected_score

elo_results <- data.frame(
  player_name = players$player_name,
  pre_rating = players$pre_rating,
  actual_score = players$total_points,
  expected_score = players$expected_score,
  difference = players$difference
)

knitr::kable(elo_results, digits = 3)
```

Rounds without an opponent number are excluded from Elo expectations. Actual scores are the reported tournament totals; any bye or forfeit points included in those totals remain included in this comparison. Positive differences indicate overperformance, and negative differences indicate underperformance. Rankings use unrounded values.

## Top Five Overperformers

```{r overperformers}
overperformers <- elo_results %>%
  arrange(desc(difference), player_name) %>%
  head(5)

knitr::kable(overperformers, digits = 3)
```

## Top Five Underperformers

```{r underperformers}
underperformers <- elo_results %>%
  arrange(difference, player_name) %>%
  head(5)

knitr::kable(underperformers, digits = 3)
```

## Findings

`r overperformers$player_name[1]` had the largest actual-minus-expected difference, scoring `r overperformers$actual_score[1]` points compared with `r round(overperformers$expected_score[1], 3)` expected points. `r underperformers$player_name[1]` had the smallest difference, scoring `r underperformers$actual_score[1]` points compared with `r round(underperformers$expected_score[1], 3)` expected points.

These differences compare tournament outcomes with expectations based on pre-tournament ratings and the actual opponents faced.
